---
title: "HMRF_Phillips"
author: "Francesco Massaini"
format: html
editor: visual
---

```{r}
library(Giotto)
library(dplyr)
library(tibble)
library(readxl)
library(data.table)
library(ggplot2)

```

```{r}
phillips = as.data.table(read_xlsx('./41467_2021_26974_MOESM3_ESM_-_Objects.xlsx'))

ggplot(phillips, aes(x = X, y = Y, color = FileName)) +
  geom_point() +
  labs(title = "Scatterplot",
       x = "X",
       y = "Y",
       color = "Filename") +
  theme_minimal()

min_max_df <- phillips[, .(
  min_X = min(X),
  max_X = max(X),
  min_Y = min(Y),
  max_Y = max(Y)
), by = FileName]




#phillips = phillips[FileName == "reg056"]
```

```{r}
# Code Block 2 - Set Paths

my_working_dir = './phillips_output/'
```

```{r}
# Code Block 3 - Set Giotto instructions

#  set Giotto instructions
instrs = createGiottoInstructions(save_plot = FALSE,
                                  show_plot = TRUE,
                                  save_dir = my_working_dir,
                                  python_path = NULL)

# Subset Phillips ####
marker_col_names = c('FOXP3', 'GATA3', 'MUC-1', 'Vimentin', 'T-bet', 'CD62L', 'Cytokeratin', 'PD-L1', 'Ki-67', 'CD15', 'CD30', 'CD2', 'GranzymeB', 'CD5', 'CD39', 'MMP9', 'CD4', 'LAG3', 'CD25', 'CD56', 'CD20', 'PD-1', 'CD11c', 'CLA-CD162', 'IDO-1', 'VISTA', 'HLA-DR', 'ICOS', 'BCL-2', 'CD3', 'CD69', 'CD8', 'CD7', 'CD45RA', 'CD45', 'CD1a', 'CD57', 'B-catenin', 'CD45RO', 'CD71', 'CD34', 'CD68', 'CD38', 'CollagenIV', 'CD31', 'Podoplanin', 'CD138', 'CD163', 'Mastcell-tryptase', 'MMP12', 'CD164', 'p53', 'CCR6', 'CD16', 'CD11b', 'CCR4', 'EGFR', 'HOECHST1:Cyc_1_ch_1', 'DRAQ5')
coord_col_names = c('X', 'Y')
meta_names = c("Count",'ClusterName', "FileName", "Patients") # FileName = sample ClusterName = cell types

# create file with offset information
my_offset_file = data.table::data.table(sample = unique(phillips$FileName),
                                        x_offset = 5000*seq(0, length(unique(phillips$FileName))),
                                        y_offset = rep(0, length(unique(phillips$FileName))))
my_offset_file$x_offset = sample(my_offset_file$x_offset)
# create a stitched image
my_offset_file <- as.data.table(my_offset_file)

phillips[my_offset_file, on = .(FileName = sample), X := X + i.x_offset]



ggplot(phillips, aes(x = X, y = Y, color = FileName)) +
  geom_point() +
  labs(title = "Scatterplot",
       x = "X",
       y = "Y",
       color = "Filename") +
  theme_minimal()

# subset phillips to create Giotto object
t_expr = as.data.frame(t(column_to_rownames(phillips[, c("Count", ..marker_col_names)], var = "Count"))) 
expr = as.data.table(rownames_to_column(t_expr, var = "markers"))
setnames(expr, names(expr), as.character(names(expr)))

coord = as.data.table(phillips[, c(..coord_col_names, "Count")]) 
colnames(coord) = c("sdimx", "sdimy", "Count") 
coord[, Count := as.character(Count)] 

meta = as.data.table(phillips[, ..meta_names])
meta[, Count := as.character(Count)] 

```

Expression file formatting

Expression count matrices accepted by Giotto should have columns correlating to individual cells (cell IDs) and rows to individual features (feature IDs).

data.table objects and text-based files ie: (.csv, .tsv, .txt) to be read in should have column 1 be feature IDs.

Locations file formatting

Numerical columns will be interpreted in order of x, y, and (optionally) z coordinate. The first non-numerical column will be taken as cell IDs

```{r}
# Code Block 4 - Create Giotto object
giotto = createGiottoObject(expression = expr,
                            expression_matrix_class = "custom",
                            expression_feat = "protein",
                            spatial_locs = coord,
                            instructions = instrs,
                            offset_file = my_offset_file)

giotto = addCellMetadata(gobject = giotto,
                         new_metadata = meta,
                         by_column = T,
                         column_cell_ID = "Count")

# visualize

cell_color_code = Giotto::getDistinctColors(length(unique(meta$ClusterName)))

spatPlot(gobject = giotto,
         cell_color = "ClusterName",
         color_as_factor = T,
         cell_color_code = cell_color_code)

giottoplot = data.frame(sdimx = giotto@spatial_locs[["cell"]][["raw"]]@coordinates[["sdimx"]], 
                        sdimy = giotto@spatial_locs[["cell"]][["raw"]]@coordinates[["sdimy"]],
                        FileName = giotto@cell_metadata[["cell"]][["protein"]]@metaDT[["FileName"]])

ggplot(giottoplot, aes(x = sdimx, y = sdimy, color = FileName)) +
  geom_point() +
  labs(title = "Scatterplot",
       x = "X",
       y = "Y",
       color = "Filename") +
  theme_minimal()
```

```{r}
giotto <- createSpatialGrid(gobject = giotto,
                                sdimx_stepsize = 500,
                                sdimy_stepsize = 500,
                                minimum_padding = 50)
spatPlot(gobject = giotto, show_grid = T, point_size = 1.5)

## delaunay network: stats + creation
plotStatDelaunayNetwork(gobject = giotto, maximum_distance = 400, save_plot = F)
giotto = createSpatialNetwork(gobject = giotto, minimum_k = 2, maximum_distance_delaunay = 400)
## create spatial networks based on k and/or distance from centroid
giotto <- createSpatialNetwork(gobject = giotto, method = 'kNN', k = 5, name = 'spatial_network')
# giotto <- createSpatialNetwork(gobject = giotto, method = 'kNN', k = 10, name = 'large_network')

# ## visualize different spatial networks on first field (~ layer 1)
# cell_metadata = pDataDT(giotto)
# field1_ids = cell_metadata[FOV == 0]$cell_ID
# subgiotto = subsetGiotto(giotto, cell_ids = field1_ids)
spatPlot2D(gobject = giotto, show_network = T,
         network_color = 'blue', spatial_network_name = 'Delaunay_network',
         point_size = 2.5, cell_color = 'ClusterName',
         cow_rel_h = 1,
         cow_rel_w = 1,
         coord_fix_ratio = 1, 
         save_plot = T)

network = giotto@spatial_network[["cell"]][["Delaunay_network"]]@networkDT

plot = ggplot() +
  geom_segment(data = network, aes(x = sdimx_begin, y = sdimy_begin, xend = sdimx_end, yend = sdimy_end), 
               color = "blue", size = 1.5, alpha = 0.8) + # Linea blu più spessa
  # Aggiungi i nodi del network
  geom_point(data = giottoplot, aes(x = sdimx, y = sdimy, color = FileName), size = 2) +
  labs(title = "Scatterplot",
       x = "X",
       y = "Y",
       color = "Filename") +
  theme_minimal()

ggsave("./phillips_output/plot_Delaunay.png", plot = plot, width = 60, height = 6, limitsize = FALSE)


```

```{r}
# giotto = normalizeGiotto(gobject = giotto, scalefactor = 6000, verbose = T)

giotto = binSpect(giotto, return_gobject = TRUE, expression_values = "raw")
 
giotto <- runPCA(gobject = giotto, feats_to_use = "all_genes", scale_unit = F, center = F, expression_values = "raw")
screePlot(giotto)
plotPCA(giotto)
 
giotto <- runUMAP(giotto, dimensions_to_use = 1:15, n_threads = 10, expression_values = "raw")
plotUMAP(gobject = giotto)
 
giotto <- runtSNE(giotto, dimensions_to_use = 1:15, expression_values = "raw")
plotTSNE(gobject = giotto)
 
 
## sNN network (default) on previously computed dim reduction 
giotto <- createNearestNetwork(
   gobject = giotto, 
   dim_reduction_to_use = "umap",  # "umap" as it is the reduction method
   spat_unit = "cell", 
   feat_type = "protein", 
   dim_reduction_name = "protein.umap",  # Exact name in your object
   k = 10
 )
 ## Leiden clustering on nn_network_to_use = "sNN" (default)
 # Perform Leiden clustering using the created sNN network
giotto <- doLeidenCluster(
   gobject = giotto,
   nn_network_to_use = "sNN",  # The network type to use
   spat_unit = "cell", 
   feat_type = "protein", 
   network_name = "sNN.umap",  # The exact name of the network created
   resolution = 0.4,
   n_iterations = 1000
 )
# 
# 
plotUMAP(gobject = giotto,
          cell_color = 'leiden_clus',
          show_NN_network = T,
          nn_network_to_use = "sNN",  # Specify the correct network type
          network_name = "sNN.umap",  # Specify the correct network name
          point_size = 2.5)

# giotto@dimension_reduction$cells$spatial <- list()
# giotto@dimension_reduction$cells$spatial$spatial_feat <- NULL
# giotto@dimension_reduction$cells$spatial$spatial_feat$name <- "spatial_feat"
# giotto@dimension_reduction$cells$spatial$spatial_feat$reduction_method <- "spatial"


HMRF_init_obj <- initHMRF_V2(gobject = giotto, cl.method = "km",
                             expression_values = "raw",
                             # use_spatial_genes = my_spatial_genes,
                             # spatial_network_name = 'Delaunay_network',
                             spat_unit = "cell", 
                             feat_type = "protein",
                             k = 10)

#' This function will run a HMRF model after initialization of HMRF. Of note
#' is the beta parameter, the smoothing parameter.
#' If the users are interested in selecting results from different smoothness,
#' we recommend running a range of betas,
#' hence betas specify what this range is. For example, betas=c(0,10,5) will
#' run for the following betas: 0, 10, 20, 30, 40.
#' betas=c(0,5,2) will run for betas: 0, 5, 10. Setting the beta can use the
#' following guideline:
#' If number of features N is 10<N<50, set betas=c(0, 1, 20)
#' For 50<N<100, set betas=c(0, 2, 25)
#' For 100<N<500, set betas=c(0, 5, 20)
#' If user didn't specify the number or sequence of beta, the function will
#' provide a default value based on the number of features.

HMRF_spatial_genes = doHMRF_V2(HMRF_init_obj = HMRF_init_obj)
# Default value beta = 6 is used...[1] "Doing beta=6.000"

save(HMRF_init_obj, file = "./HMRF_init_obj_leiden.RData")
save(HMRF_spatial_genes, file = "./HMRF_spatial_genes_leiden.RData")
# save(giotto, file = "./giotto.RData")

library(tibble)
library(dplyr)
library(ggplot2)
df = as.data.frame(HMRF_spatial_genes[[1]]) 
df = rownames_to_column(df, var = "cell_ID")

coord = giotto@spatial_locs[["cell"]][["raw"]]@coordinates

merge = left_join(df, coord)

niches = ggplot(data = merge, aes(x = sdimx, y = sdimy, color =  as.character(class) )) +
  geom_point() +
  theme_minimal()

ggsave("./phillips_output/niches_k_means_shuffled_sample.png", plot = niches, width = 100, height = 6, limitsize = FALSE)


```

```{r}
CellClus <- merge[,c("cell_ID","class")]
CellFile <- meta[,c("FileName","Count", "Patients", "ClusterName")]
colnames(CellFile) <- c("FileName","cell_ID", "Patients", "cell_type")

final_df = merge(CellClus, CellFile)
save(final_df, file = "Phillips_HMRF_niches.RData")

ClusterCount <- as.data.frame(table(merge(CellClus, CellFile ) [, c("FileName","class")] ))

M <- as.matrix(table(merge(CellClus, CellFile ) [, c("FileName","class")] ))
pheatmap::pheatmap(M)
```
